---
title: "Lab3 Crime Statistics"
author: "Kenneth Chen, Peter Trenkwalder, Danielle Salah"
date: "7/15/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
library(car)
```

```{r}
crime = read.csv("crime_v2.csv")
```

# Introduction

We received a crime dataset on North Carolina and would like to explore crime statistics. We like to investigate crime statistics at hands to develop several viable approaches in order to reduce crime. The dataset has  **`r nrow(crime)`** observations and **`r ncol(crime)`** variables. Our first approach is to investigate each of the variables and how they relate to the occurence of crimes. 

# Exploratory Data Analysis

```{r echo=FALSE, include=FALSE}
# Not displaying the output because it's the same as my writeup
ls.str(pat='crime')
```

We listed all variables and their descriptions here. 

| variable | label |
|--------|-------|
| 1 county    | county identifier     |
| 2 year   | 1987     |
| 3 crmrte   | crimes committed per person     |
| 4 prbarr   | 'probability' of arrest     |
| 5 prbconv    | 'probability' of conviction     |
| 6 prbpris   | 'probability' of prison sentence     |
| 7 avgsen   | avg. sentence, days     |
| 8 polpc    | police per capita     |
| 9 density  | people per sq. mile |
| 10 taxpc   | tax revenue per capita |
| 11 west    | =1 if in western N.C. |
| 12 central | =1 if in central N.C. |
| 13 urban   | =1 if in SMSA |
| 14 pctmin80 | perc. minority, 1980 |
| 15 wcon    | weekly wage, construction |
| 16 wtuc    | weekly wage, trns, util, commun |
| 17 wtrd    | weekly wage, whlelse, retail trade |
| 18 wfir    | weekly wage, fin, ins, real est |
| 19 wser    | weekly wage, service industry |
| 20 wmfg    | weekly wage, manufacturing |
| 21 wfed    | weekly wage, fed employees |
| 22 wsta    | weekly wage, state employees |
| 23 wloc    | weekly wage, local gov emps |
| 24 mix     | offense mix: face-to-face/other |
| 25 pctymle | percent young male |

Out of 25 variables, we set our dependent variable to be **crime rates, crmrte** because we believe this reflects the frequency of crimes in North Carolina. 

# Questions      
**1. Are crime rates affected by probability of arrest? or can we predict the crime rate by the probability of arrest in a given location?**  

Answer

**2. Are crime rates dependent on the population of young males?**   


$\widehat{\text{crime rate}} = \beta_0 + \beta_1 \cdot \text{prbarr} + \beta_2 \cdot \text{pctyml}$


**3. Omitted Variable Bias (OVB)**   

We are concerned that the variable we are currently interested, probability of arrest and population of young males, have other variables that are highly corrrelated to each other such as the location in North Carolia, and if the wages across the country has a causal effect on crimes based on monetary incentive. If that's the case, we will be missing a key variable in our analysis. 


## Preliminary investigation of each variable correlation 

In order to retrieve all observations equally across all 25 variables, we first filtered the data with complete data on all 25 variables. Subsequently we checked each of the variable relationship by scatterplot matrix 5 at a time. In the last plot, we focused on 

1. crime rates, 
2. probabiilty of arrest, 
3. police per capita, 
4. weekly wages, federal employee, 
5. weekly wages, state, 
6. percent of young males 

```{r}
# Since we're going to look at the all the variables across, I'm dropping all the empty roles here.
# So 97 observations end up at 91 observations. 
crime_fullrow = crime[complete.cases(crime), ]
```

# Checking the first 5 variables
```{r}
scatterplotMatrix(crime_fullrow[1:5])
```

# Checking the next 5 variables
```{r}
scatterplotMatrix(crime_fullrow[6:10])
```

# Checking the next 5 variables
```{r}
scatterplotMatrix(crime_fullrow[11:15])
```

# Checking the next 5 variables
```{r}
scatterplotMatrix(crime_fullrow[16:20])
```

# Checking the last 5 variables
```{r}
scatterplotMatrix(crime_fullrow[21:25])
```

# Looking at the crime rate, probability arrest and a few other variables or (Key Variables) 
```{r}
scatterplotMatrix(~ crmrte + prbarr + polpc + wfed + wsta + pctymle + pctmin80, data=crime_fullrow)
```

To use the stargazer later on
#stargazer
{r, results='asis'}
library(stargazer)
stargazer(model1, model2, type = "latex", 
          report = "vc", # Don't report errors, since we haven't covered them
          title = "Linear Models Predicting College GPA",
          keep.stat = c("rsq", "n"),
          omit.table.layout = "n") # Omit more output related to errors
          
          




# Police

```{r}
hist(crime$polpc,
     main = "Histogram of Population of Police",
     xlab = 'Police per Capita',
     ylab = 'Frequency',
     col = 'grey')

```

As evidenced in the distribution above, the majority of the counties represented in the data set have 0.001 to 0.002 police members per capita. The other notable feature of this visual is the county with a police per capita that is a noticable outlier at 0.009 to 0.01 police members per capita. We will first investigate these two cases.

```{r}

polpc_low <- crime$county[crime$polpc > 0.001 & crime$polpc < 0.002]
polpc_high <- crime$county[crime$polpc > 0.009]

polpc_high

view(crime$county == 115)

```

*** add more here

An additionally helpful analyis is to look at how the rate of police per capita varies by size of the county.

```{r}

scatterplot(crime$density, crime$polpc,
            main = 'Police per Capita vs Population Density',
            xlab = 'People per Square Mile',
            ylab = 'Police per Capita')

```


While police per capital is clearly already a normalized value, this view helps illustrate that more dense regions require more police per capita, and more rural areas require fewer.

```{r}

urb_pol <- crime$polpc[crime$urban ==1]

scatterplot(urb_pol, crime$density[crime$urban==1])


```

```{r}

plot(lm(crmrte ~ polpc, data = crime))

performanceanalytics
chart.correlation
```

#Insights

larger cities: denser population, more police, more crime
Policies: 